Imputation of Incomplete Multilevel Data with \pkg{mice}

### 2022-03-07 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
Feedback md pattern plot

@@

This papers serves as a tutorial for imputing incomplete multilevel data with \pkg{mice}. We provide practical guidelines and code snippets for different missing data situations, including missing not at random (MNAR) mechanisms [where the probability to be missing depends on unrecorded information, making the missingness non-ignorable, @rubi76; @meng94]. For reasons of brevity, we focus on imputation by chained equations wit \pkg{mice} exclusively^[Note that the alternative, joint modeling imputation for multilevel data or \pkg{jomo} @jomo, has been implemented in \pkg{mice} as well but is outside the scope of this tutorial.]. Other useful resources for the analysis of incomplete multilevel data include the \proglang{R} packages \pkg{mitml}, \pkg{miceadds}, and \pkg{mdmb}, and empirical work by @audi18 and @grun18. Please note that this tutorial paper assumes a basic level of knowledge on multilevel models.^[Note to self: We're providing an overview of implementations. It's up-to the reader to decide which multilevel strategy suits their data. We won't go into detail for the different methods (and equations). This paper is just a software tutorial, so we'll keep it practical.] Assumed knowledge also includes the use of the 'piping operator', `%>%`, adopted from the \pkg{magrittr} package, and the \pkg{lme4} notation for multilevel models.^[Add environment info, seed and version number(s) somewhere!]

@@

For each of these datasets, we will discuss the nature of the missingness, choose one or more imputation models and evaluate the imputed data, but we will also highlight one specific aspect of the imputation workflow. With the `popmis` data, we show how (and how not) to develop an imputation model. With the `hiv` data we focus on extending the imputation model to include Heckman-type selection-inclusion methods. With the `impact` data we provide an example of multivariate missingness in real-world data. Together, this should give enough scaffolding for applied researchers who are faced with incomplete multilevel data.

@@

```{r comp_est, echo=FALSE}
est_comp_fixed %>% setNames(c("", "Estimate with 95% CI"))
```

@@

```{r pop_norm}
pred <- ini$pred
pred["popular", ] <- c(-2, 2, 2, 2) 
#-2 for the cluster variable, 2 for random effects
meth <- ini$meth
meth <- c("", "2l.norm", "",  "")
# meth <- c("", "2l.pmm", "",  "")
imp_norm_2l <-
  mice(
    popmis %>% mutate(school = as.integer(school)),
    pred = pred,
    meth = meth,
    maxit = 1,
    print = FALSE
  )
```

```{r pop_norm_eval, echo=FALSE}
# plot(imp_norm)
plot_imps(imp_norm_2l, x = "popular", type = "stripplot")

@@

  - `name` Name of the study,
  - `type` Type of study (RCT: randomized controlled trial, OBS: observational cohort),
  - `age` Age of the patient,
  - `motor_score` Glasgow Coma Scale motor score,
  - `pupil` Pupillary reactivity,
  - `ct` Marshall Computerized Tomography classification,
  <!-- - `hypox` Hypoxia (0=no, 1=yes), -->
  <!-- - `hypots` Hypotension (0=no, 1=yes), -->
  <!-- - `tsah` Traumatic subarachnoid hemorrhage (0=no, 1=yes), -->
  <!-- - `edh` Epidural hematoma (0=no, 1=yes), -->
  - `mort` 6-month mortality (0=alive, 1=dead).

@@

`glmer(mort ~ 1 + type + age + motor_score + pupil + ct + hypox + hypots + tsah + edh + (1 | name))`

@@

```{r impact, echo=FALSE, fig.width=6, fig.height=8}
# load data
data("impact", package = "metamisc")
impact_NA <- readRDS("../Data/impact_NA.RDS") 
# # descriptive statistics
# by(impact, impact$name, summary) 
# psych::describe(impact)[,c(2:5,8:9)]
# # missingness
plot_md_pat(impact_NA)
# TODO: rotate variable labels
# TODO: think about nr of clusters: if there's too many, faceting is no option -> some distribution across the clusters like shepley plot
# TODO: make missingness % gradient for clusters (ie the number of clusters for which this variable is missing).
```

@@

```{r impact_cca}
mod_cca <- glmer(mort ~ type + age + motor_score + pupil + ct + (1 | name), family = "binomial", data = impact_NA) 
mod_cca %>% broom.mixed::tidy()

@@

>>>>

<!-- Missing data pattern with clustering -->
<!-- TODO: simplify argument in md pattern plot and/or network plot for md patterns (summarized per variable?) -->
<!-- TODO: test if there's a random distribution of missingness in the md pattern? -->
<!-- TODO: check spss missing data pie chart? -->
<!-- TODO: similarity of patterns over clusters -->
<!-- TODO: gradient (alpha) to show in how many of the clusters the row in the pattern occurs -->
<!-- TODO: 3 columns could be 3 boxplots, over these clusters we have a distribution of these statistics across clusters.  -->
<!-- TODO: look at vim -->

@@

This papers serves as a tutorial for imputing incomplete multilevel data with \pkg{mice}. We provide practical guidelines and code snippets for different missing data situations, including missing not at random (MNAR) mechanisms [where the probability to be missing depends on unrecorded information, making the missingness non-ignorable, @rubi76; @meng94]. For reasons of brevity, we focus on imputation by chained equations wit \pkg{mice} exclusively^[Note that the alternative, joint modeling imputation for multilevel data or \pkg{jomo} @jomo, has been implemented in \pkg{mice} as well but is outside the scope of this tutorial.]. Other useful resources for the analysis of incomplete multilevel data include the \proglang{R} packages \pkg{mitml}, \pkg{miceadds}, and \pkg{mdmb}, and empirical work by @audi18 and @grun18. Please note that this tutorial paper assumes a basic level of knowledge on multilevel models.^[Note to self: We're providing an overview of implementations. It's up-to the reader to decide which multilevel strategy suits their data. We won't go into detail for the different methods (and equations). This paper is just a software tutorial, so we'll keep it practical.] Assumed knowledge also includes the use of the 'piping operator', `%>%`, adopted from the \pkg{magrittr} package, and the \pkg{lme4} notation for multilevel models.^[TODO: Add environment info, seed and version number(s) somewhere!]

@@

For each of these datasets, we will discuss the nature of the missingness, choose one or more imputation models and evaluate the imputed data, but we will also highlight one specific aspect of the imputation workflow. With the `popmis` data, we show how (and how not) to develop an imputation model. With the `hiv` data we focus on extending the imputation model to include Heckman-type selection-inclusion methods. With the `impact` data we provide an example of multivariate missingness in real-world data. Together, this should give enough scaffolding for applied researchers who are faced with incomplete multilevel data.^[TODO: Add notation paragraph, linking multilevel equations to \pkg{lme4} formulas.]

@@

```{r comp_est, echo=FALSE}
est_comp_fixed %>% setNames(c("", "Estimate with 95% CI"))
# TODO: make this a figure with the different methods as facets
```

@@

```{r pop_norm}
pred <- ini$pred
pred["popular", ] <- c(-2, 2, 2, 2) 
#-2 for the cluster variable, 2 for random effects
meth <- ini$meth
meth["popular"] <- "2l.pmm"
imp_pmm_2l <-
  mice(
    popmis %>% mutate(school = as.integer(school)),
    pred = pred,
    meth = meth,
    maxit = 1,
    print = FALSE
  )
```

```{r pop_norm_eval, echo=FALSE}
# plot(imp_norm)
plot_imps(imp_pmm_2l, x = "popular", type = "stripplot")

@@

  - `name` Name of the study,
  - `type` Type of study (RCT: randomized controlled trial, OBS: observational cohort),
  - `age` Age of the patient,
  - `motor_score` Glasgow Coma Scale motor score,
  - `pupil` Pupillary reactivity,
  - `ct` Marshall Computerized Tomography classification,
  - `hypox` Hypoxia (0=no, 1=yes),
  - `hypots` Hypotension (0=no, 1=yes),
  - `tsah` Traumatic subarachnoid hemorrhage (0=no, 1=yes),
  - `edh` Epidural hematoma (0=no, 1=yes),
  - `mort` 6-month mortality (0=alive, 1=dead).

@@

`glmer(mort ~ 1 + type + age + motor_score + pupil + ct + (1 | name))`

<!-- + hypox + hypots + tsah + edh  -->

@@

```{r impact, echo=FALSE, fig.width=6, fig.height=8}
# load data
data("impact", package = "metamisc")
impact_NA <- readRDS("../Data/impact_NA.RDS") #%>% select(!c(hypox, hypots, tsah, edh))

# # descriptive statistics
# by(impact, impact$name, summary) 
# psych::describe(impact)[,c(2:5,8:9)]
# # missingness
plot_md_pat(impact_NA) 
# TODO: rotate variable labels
# TODO: think about nr of clusters: if there's too many, faceting is no option -> some distribution across the clusters like shepley plot
# TODO: make missingness % gradient for clusters (ie the number of clusters for which this variable is missing).
```

@@

```{r impact_cca, echo=FALSE}
# mod_cca <- glmer(mort ~ type + age + motor_score + pupil + ct + (1 | name), family = "binomial", data = impact_NA) 
mod_cca_dummy <- glm(mort ~ type + age + motor_score + pupil + ct + as.factor(name), family = "binomial", data = impact_NA) 
mod_cca_dummy %>% broom.mixed::tidy()

@@

If we run the multilevel model on the incomplete date, we get a warning about unindentifyability. This means that with CCA as missing data method, we cannot trust the estimates. To still obtain some estimates to compare later, we fit a simplified analysis model: using the cluster variable as indicator in the model.

### 2022-03-07 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex
Feedback md pattern plot

This papers serves as a tutorial for imputing incomplete multilevel data
with \pkg{mice}. We provide practical guidelines and code snippets for
different missing data situations, including missing not at random
(MNAR) mechanisms \citep[where the probability to be missing depends on
unrecorded information, making the missingness
non-ignorable,][]{rubi76, meng94}. For reasons of brevity, we focus on
imputation by chained equations wit \pkg{mice} exclusively\footnote{Note
  that the alternative, joint modeling imputation for multilevel data or
  \pkg{jomo} \citet{jomo}, has been implemented in \pkg{mice} as well
  but is outside the scope of this tutorial.}. Other useful resources
for the analysis of incomplete multilevel data include the \proglang{R}
packages \pkg{mitml}, \pkg{miceadds}, and \pkg{mdmb}, and empirical work
by \citet{audi18} and \citet{grun18}. Please note that this tutorial
paper assumes a basic level of knowledge on multilevel
models.\footnote{Note to self: We're providing an overview of
  implementations. It's up-to the reader to decide which multilevel
  strategy suits their data. We won't go into detail for the different
  methods (and equations). This paper is just a software tutorial, so
  we'll keep it practical.} Assumed knowledge also includes the use of
the `piping operator', \texttt{\%\textgreater{}\%}, adopted from the
\pkg{magrittr} package, and the \pkg{lme4} notation for multilevel
models.\footnote{Add environment info, seed and version number(s)
  somewhere!}

@@

For each of these datasets, we will discuss the nature of the
missingness, choose one or more imputation models and evaluate the
imputed data, but we will also highlight one specific aspect of the
imputation workflow. With the \texttt{popmis} data, we show how (and how
not) to develop an imputation model. With the \texttt{hiv} data we focus
on extending the imputation model to include Heckman-type
selection-inclusion methods. With the \texttt{impact} data we provide an
example of multivariate missingness in real-world data. Together, this
should give enough scaffolding for applied researchers who are faced
with incomplete multilevel data.

@@

\begin{CodeChunk}
\begin{CodeInput}
R> pred <- ini$pred
R> pred["popular", ] <- c(-2, 2, 2, 2) 
R> #-2 for the cluster variable, 2 for random effects
R> meth <- ini$meth
R> meth <- c("", "2l.norm", "",  "")
R> # meth <- c("", "2l.pmm", "",  "")
R> imp_norm_2l <-
+   mice(
+     popmis %>% mutate(school = as.integer(school)),
+     pred = pred,
+     meth = meth,
+     maxit = 1,
+     print = FALSE
+   )
\end{CodeInput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeOutput}
Warning: Removed 848 rows containing non-finite values (stat_boxplot).
\end{CodeOutput}
\begin{CodeOutput}
Warning: Removed 848 rows containing missing values (geom_point).
\end{CodeOutput}

@@

\begin{CodeChunk}
\begin{CodeOutput}
Warning: Removed 848 rows containing non-finite values (stat_boxplot).
\end{CodeOutput}
\begin{CodeOutput}
Warning: Removed 848 rows containing missing values (geom_point).
\end{CodeOutput}

@@

\begin{itemize}
\tightlist
\item
  \texttt{name} Name of the study,
\item
  \texttt{type} Type of study (RCT: randomized controlled trial, OBS:
  observational cohort),
\item
  \texttt{age} Age of the patient,
\item
  \texttt{motor\_score} Glasgow Coma Scale motor score,
\item
  \texttt{pupil} Pupillary reactivity,
\item
  \texttt{ct} Marshall Computerized Tomography classification,
\item
  \texttt{mort} 6-month mortality (0=alive, 1=dead).
\end{itemize}

@@

\texttt{glmer(mort\ \textasciitilde{}\ 1\ +\ type\ +\ age\ +\ motor\_score\ +\ pupil\ +\ ct\ +\ hypox\ +\ hypots\ +\ tsah\ +\ edh\ +\ (1\ \textbar{}\ name))}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> mod_cca <- glmer(mort ~ type + age + motor_score + pupil + ct + (1 | name), family = "binomial", data = impact_NA) 
\end{CodeInput}
\begin{CodeOutput}
Warning in checkConv(attr(opt, "derivs"), opt$par, ctrl = control$checkConv, : Model is nearly unidentifiable: very large eigenvalue
 - Rescale variables?
\end{CodeOutput}
\begin{CodeInput}
R> mod_cca %>% broom.mixed::tidy()
\end{CodeInput}
\begin{CodeOutput}
# A tibble: 9 x 7
  effect   group term            estimate std.error statistic   p.value
  <chr>    <chr> <chr>              <dbl>     <dbl>     <dbl>     <dbl>
1 fixed    <NA>  (Intercept)      -2.66     0.194      -13.7   1.11e-42
2 fixed    <NA>  typeRCT          -0.412    0.172       -2.40  1.64e- 2
3 fixed    <NA>  age               0.0298   0.00190     15.7   2.56e-55
4 fixed    <NA>  motor_score3     -0.631    0.0823      -7.66  1.82e-14
5 fixed    <NA>  motor_score4     -0.994    0.0845     -11.8   5.61e-32
6 fixed    <NA>  motor_score5/6   -1.54     0.0868     -17.7   2.88e-70
7 fixed    <NA>  pupil             0.438    0.0397      11.0   3.01e-28
8 fixed    <NA>  ct                0.430    0.0343      12.5   5.34e-36
9 ran_pars name  sd__(Intercept)   0.257   NA           NA    NA       
\end{CodeOutput}
\begin{CodeInput}
R> # use glm with name as dummy
R> # only random effect on ct, not the other variables
R> # make forest plot of ct variable
R> # is there heterogen in association of ct with mort?
R> # is there any random effect on that variable at all?
R> # 
\end{CodeInput}
\end{CodeChunk}

>>>>

This papers serves as a tutorial for imputing incomplete multilevel data
with \pkg{mice}. We provide practical guidelines and code snippets for
different missing data situations, including missing not at random
(MNAR) mechanisms \citep[where the probability to be missing depends on
unrecorded information, making the missingness
non-ignorable,][]{rubi76, meng94}. For reasons of brevity, we focus on
imputation by chained equations wit \pkg{mice} exclusively\footnote{Note
  that the alternative, joint modeling imputation for multilevel data or
  \pkg{jomo} \citet{jomo}, has been implemented in \pkg{mice} as well
  but is outside the scope of this tutorial.}. Other useful resources
for the analysis of incomplete multilevel data include the \proglang{R}
packages \pkg{mitml}, \pkg{miceadds}, and \pkg{mdmb}, and empirical work
by \citet{audi18} and \citet{grun18}. Please note that this tutorial
paper assumes a basic level of knowledge on multilevel
models.\footnote{Note to self: We're providing an overview of
  implementations. It's up-to the reader to decide which multilevel
  strategy suits their data. We won't go into detail for the different
  methods (and equations). This paper is just a software tutorial, so
  we'll keep it practical.} Assumed knowledge also includes the use of
the `piping operator', \texttt{\%\textgreater{}\%}, adopted from the
\pkg{magrittr} package, and the \pkg{lme4} notation for multilevel
models.\footnote{TODO: Add environment info, seed and version number(s)
  somewhere!}

@@

For each of these datasets, we will discuss the nature of the
missingness, choose one or more imputation models and evaluate the
imputed data, but we will also highlight one specific aspect of the
imputation workflow. With the \texttt{popmis} data, we show how (and how
not) to develop an imputation model. With the \texttt{hiv} data we focus
on extending the imputation model to include Heckman-type
selection-inclusion methods. With the \texttt{impact} data we provide an
example of multivariate missingness in real-world data. Together, this
should give enough scaffolding for applied researchers who are faced
with incomplete multilevel data.\footnote{TODO: Add notation paragraph,
  linking multilevel equations to \pkg{lme4} formulas.}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> pred <- ini$pred
R> pred["popular", ] <- c(-2, 2, 2, 2) 
R> #-2 for the cluster variable, 2 for random effects
R> meth <- ini$meth
R> meth["popular"] <- "2l.pmm"
R> imp_pmm_2l <-
+   mice(
+     popmis %>% mutate(school = as.integer(school)),
+     pred = pred,
+     meth = meth,
+     maxit = 1,
+     print = FALSE
+   )
\end{CodeInput}
\end{CodeChunk}

@@

@@

\begin{itemize}
\tightlist
\item
  \texttt{name} Name of the study,
\item
  \texttt{type} Type of study (RCT: randomized controlled trial, OBS:
  observational cohort),
\item
  \texttt{age} Age of the patient,
\item
  \texttt{motor\_score} Glasgow Coma Scale motor score,
\item
  \texttt{pupil} Pupillary reactivity,
\item
  \texttt{ct} Marshall Computerized Tomography classification,
\item
  \texttt{hypox} Hypoxia (0=no, 1=yes),
\item
  \texttt{hypots} Hypotension (0=no, 1=yes),
\item
  \texttt{tsah} Traumatic subarachnoid hemorrhage (0=no, 1=yes),
\item
  \texttt{edh} Epidural hematoma (0=no, 1=yes),
\item
  \texttt{mort} 6-month mortality (0=alive, 1=dead).
\end{itemize}

@@

\texttt{glmer(mort\ \textasciitilde{}\ 1\ +\ type\ +\ age\ +\ motor\_score\ +\ pupil\ +\ ct\ +\ (1\ \textbar{}\ name))}

@@

\begin{CodeChunk}
\begin{CodeOutput}
# A tibble: 22 x 5
   term                 estimate std.error statistic  p.value
   <chr>                   <dbl>     <dbl>     <dbl>    <dbl>
 1 (Intercept)           -3.37     0.205     -16.4   1.97e-60
 2 typeRCT                0.395    0.177       2.23  2.59e- 2
 3 age                    0.0301   0.00191    15.8   5.01e-56
 4 motor_score3          -0.618    0.0829     -7.45  9.33e-14
 5 motor_score4          -1.00     0.0850    -11.8   6.07e-32
 6 motor_score5/6        -1.53     0.0874    -17.5   7.92e-69
 7 pupil                  0.434    0.0398     10.9   9.36e-28
 8 ct                     0.428    0.0344     12.4   1.50e-35
 9 as.factor(name)CSTAT  -0.177    0.199      -0.892 3.72e- 1
10 as.factor(name)EBIC    0.560    0.174       3.22  1.30e- 3
# ... with 12 more rows
\end{CodeOutput}
\end{CodeChunk}

If we run the multilevel model on the incomplete date, we get a warning
about unindentifyability. This means that with CCA as missing data
method, we cannot trust the estimates. To still obtain some estimates to
compare later, we fit a simplified analysis model: using the cluster
variable as indicator in the model.
